---
title: "HW 02"
author: "Nathan Herling"
date: "2025-06-13"
_due: "Friday-June-13-2025"   
format: 
  html:
    embed-resources: true
toc: true
#custom field '_due', 
#if you want to show in the document:**Due Date:** {{_due}}
---

## 0 - Setup

```{r setup, echo=FALSE}
#--------------------->
#################
# Package Setup #
#################
#Check if pacman [package manager] is installed, if not install it.
#throw [FYI] alert either way.
if (!requireNamespace("pacman", quietly = TRUE)) {
  message("Installing 'pacman' (not found locally)...")
  install.packages("pacman")
} else {
  message("[FYI]\n'pacman' already installed — skipping install.")
}
# use this line for installing/loading
# pacman::p_load()
# - packages to load stored in a variable (vector)
pkgs <- c(
  "tidyverse",
  "glue",
  "scales",
  "lubridate",
  "patchwork",
  "ggh4x",
  "ggrepel",
  "openintro",
  "ggridges",
  "dsbox",
  "janitor",
  "here",
  "knitr",
  "ggthemes",
  "ggplot2",
  "kableExtra"
)

# - load from the character array/vector
pacman::p_load(char=pkgs)

# - install tidyverse/dsbox directly from Git Hub
# - this allows for the possible need to install on a repo. pull.
# - and, if it's already installed just thorw an alert.
if (!requireNamespace("dsbox", quietly = TRUE)) {
  message("Installing 'dsbox' from GitHub (not found locally)...")
  suppressMessages(devtools::install_github("tidyverse/dsbox"))
} else {
  message("[FYI]\n'dsbox' already installed — skipping GitHub install.")
}

# - alert to user packages loaded.
# Set number of columns (adjustable)
n_cols <- 4

# Add * to each package name
pkgs <- paste0("* ", pkgs)

# Calculate number of rows based on total packages
n_rows <- ceiling(length(pkgs) / n_cols)

# Pad with empty strings to complete grid
pkgs_padded <- c(pkgs, rep("", n_rows * n_cols - length(pkgs)))

# Create matrix (fill by row)
pkg_matrix <- matrix(pkgs_padded, nrow = n_rows, byrow = TRUE)

# Print header
cat("The packages loaded:")

# Loop and print each row (use invisible to suppress NULL)
invisible(apply(pkg_matrix, 1, function(row) {
  cat(paste(format(row, width = 22), collapse = ""), "\n")
}))
#-------------------------->
######################
# Basic set Theme up #
######################
# ---- set theme for ggplot2
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 14))

# set width of code output
options(width = 65)

# set figure parameters for knitr
knitr::opts_chunk$set(
  fig.width = 7,        # 7" width
  fig.asp = 0.618,      # the golden ratio
  fig.retina = 3,       # dpi multiplier for displaying HTML output on retina
  fig.align = "center", # center align figures
  dpi = 300             # higher dpi, sharper image
)
## ---- end theme set up
```
## 1 - A new day, a new plot, a new geom

> **Question #1**
> <div style="background-color:#e0e0e0; padding:10px; border-radius:5px; border:1px solid #16DB65; overflow-wrap:anywhere; word-break:break-word; white-space:normal;">
> A new day, a new plot, a new geom. The goal of this exercise is to learn about a new type of plot (ridgeline plot) and to learn how to make it. Use the 
> ```geom_density_ridges()``` function from the ```ggridges``` package to make a ridge plot of Airbnb review scores of Edinburgh neighborhoods. The neighborhoods should be ordered by their median review scores. The data can be found in the ```dsbox``` package, and it’s called ```edibnb```. Also include an interpretation for your visualization. You should review feedback from your Homework 1 to make sure you capture anything you may have missed previously.
> </div>


```{r}
#| label: label-me-1
#| echo: false

# =======================================
# Diagnostics Summary Table Preparation
# =======================================

# Data Type
data_type <- class(edibnb$review_scores_rating)

# - info on the DSet
#print(help("edibnb"))

# - Summary Stats
summary_stats <- summary(edibnb$review_scores_rating)

# - Missing values
missing_count <- sum(is.na(edibnb$review_scores_rating))

# - IQR and Outlier Detection
q1 <- quantile(edibnb$review_scores_rating, 0.25, na.rm = TRUE)
q3 <- quantile(edibnb$review_scores_rating, 0.75, na.rm = TRUE)
iqr <- q3 - q1
lower_bound <- q1 - 1.5 * iqr
upper_bound <- q3 + 1.5 * iqr

outliers <- edibnb$review_scores_rating[!is.na(edibnb$review_scores_rating) &
                                          (edibnb$review_scores_rating < lower_bound |
                                           edibnb$review_scores_rating > upper_bound)]

# - Construct a table of key diagnostics
diagnostics_table <- tibble::tibble(
  Metric = c(
    "Data Type",
    "Min", "1st Quartile", "Median", "Mean", "3rd Quartile", "Max",
    "Missing Values",
    "IQR", "Lower Outlier Bound", "Upper Outlier Bound", "Outlier Count"
  ),
  Value = c(
    data_type,
    summary_stats[["Min."]],
    summary_stats[["1st Qu."]],
    summary_stats[["Median"]],
    summary_stats[["Mean"]],
    summary_stats[["3rd Qu."]],
    summary_stats[["Max."]],
    missing_count,
    iqr,
    lower_bound,
    upper_bound,
    length(outliers)
  )
)

# Print the diagnostics table with styled caption and highlighted row
diagnostics_table |>
  kable(
    caption = "<span style='font-weight:bold; font-size:1.2em;'>Table 1. Diagnostic Summary for <code>review_scores_rating</code> (edibnb data set)</span>",
    escape = FALSE
  ) |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) |>
  row_spec(
    which(diagnostics_table$Metric == "Missing Values"),   # 🔧 Highlight "Missing Values" row
    background = "#fff9c4"  # 🔧 Light yellow
  )

# =======================================
# Visualization: Ridge Plot by Neighborhood
# =======================================

# - Filter out rows with NA in neighbourhood or review_scores_rating
edibnb_clean <- edibnb |>
  filter(!is.na(neighbourhood), !is.na(review_scores_rating))

# - Calculate median review scores by neighborhood
median_scores <- edibnb_clean |>
  group_by(neighbourhood) |>
  summarize(median_score = median(review_scores_rating, na.rm = TRUE)) |>
  arrange(median_score)

# - Reorder neighborhoods by median score
edibnb_clean <- edibnb_clean |>
  mutate(neighbourhood = factor(neighbourhood, levels = median_scores$neighbourhood))

# ==== New: Calculate mean review scores for each neighborhood (for annotation) ====
mean_scores <- edibnb_clean |>
  group_by(neighbourhood) |>
  summarize(mean_score = mean(review_scores_rating, na.rm = TRUE))

# ==== Modified plot: add geom_text for mean scores offset to the right ====
# Create a named vector of colors for neighborhoods
neighborhood_colors <- setNames(viridis::viridis(length(levels(edibnb_clean$neighbourhood))), levels(edibnb_clean$neighbourhood))

# - Get the first and last neighborhood (lowest & highest median)
low_neigh <- levels(edibnb_clean$neighbourhood)[1]
high_neigh <- levels(edibnb_clean$neighbourhood)[length(levels(edibnb_clean$neighbourhood))]

# - Maximum x-position for annotation
max_annot_x <- max(mean_scores$mean_score, na.rm = TRUE) + 25

#===============
# Make the plot
#===============
g1 <- ggplot(edibnb_clean, aes(x = review_scores_rating, y = neighbourhood)) +
  geom_density_ridges(
    aes(color = neighbourhood),
    fill = "cornsilk4",
    alpha = 0.7,
    scale = 1.2,
    show.legend = FALSE
  ) +
  geom_point(
    data = mean_scores,
    aes(x = mean_score, y = neighbourhood, color = "Mean Score"),
    size = 3,
    shape = 18
  ) +
  geom_segment(
    data = mean_scores,
    aes(
      x = mean_score,
      xend = mean_score + 20,
      y = neighbourhood,
      yend = as.numeric(neighbourhood) + 0.5
    ),
    color = "black",
    linewidth = 0.3
  ) +
  ggrepel::geom_text_repel(
    data = mean_scores,
    aes(
      x = mean_score + 20,
      y = as.numeric(neighbourhood) + 0.5,
      label = round(mean_score, 1)
    ),
    hjust = 0,
    size = 3.2,
    color = "black",
    nudge_y = 0.15,
    segment.color = NA,
    direction = "y",
    box.padding = 0.3,
    point.padding = 0.5,
    show.legend = FALSE
  ) +
  annotate(             # - put an annotation next to the 'low' value
    "text",
    x = max_annot_x,
    y = 1.5,
    label = "Low",
    hjust = -1,
    size = 5,
    color = neighborhood_colors[[low_neigh]]
  ) +
  annotate(             # - put an annotation next to the 'high' value
    "text",
    x = max_annot_x,
    y = length(levels(edibnb_clean$neighbourhood)) + 0.5,
    label = "High",
    hjust = -1,
    size = 5,
    color = neighborhood_colors[[high_neigh]]
  ) +
  scale_x_continuous(name = "Review Scores Rating", limits = c(0, 150)) +
  scale_color_manual(
    name = "Legend",
    values = c(neighborhood_colors, "Mean Score" = "black"),
    breaks = "Mean Score",
    labels = "Mean Score"
  ) +
  labs(
    title = "Distribution of Airbnb Review Scores by Edinburgh Neighborhood\n", 
    y = "Neighborhood",
    caption = "Source: Inside Airbnb via Kaggle (data from June 2019)") +
  theme_ridges() +
  theme(
    legend.position = "right",
    legend.title = element_text(hjust = 0.6),
    axis.title.y = element_text(vjust = 1.5, hjust = 0.5),
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(hjust = 0.5),
    plot.caption = element_text(              # - stylizing the caption
      hjust = 0.5,    # right align
      color = "grey50",  # lighter grey
      size = 8
     )
  ) 

# - show the graph
g1

```
<div style="background-color:#e0e0e0; padding:10px; border-radius:5px; border:1px solid #16DB65; display:flex; align-items:center; justify-content:center; height:auto;">
  <strong><br>Interpretation</strong>
  The graph (`Distribution of Airbnb Review Scores by Edinburgh Neighborhood`) displays the distribution of Airbnb review scores across Edinburgh neighborhoods using ridgeline plots, with each neighborhood’s mean score marked by a diamond. The mean review scores are generally high, ranging from about 93.9 to 95.9. Some neighborhoods, like Morningside and Bruntsfield, show slightly higher average scores. The variation in score spread highlights differences in review consistency between neighborhoods, making it easier to compare where listings tend to receive better feedback.
</div>
## 2 - Foreign Connected PACs
<div style="background-color:#f0f0f0; padding:10px; border-radius:5px; border:1px solid #87CEEB;">
<b>Question #2</b><br>
<div style="background-color:#e0e0e0; padding:10px; border-radius:5px; border:1px solid #16DB65; display:flex; align-items:center; justify-content:center; height:60px;">
  <strong><br>Put the question here</strong>
</div>
</div>
```{r}
#| label: label-me-2
#| echo: false
#| fig-width: 10


# Suppress readr column type messages
options(readr.show_col_types = FALSE) 
# Load and combine data
list_of_files <- fs::dir_ls(path = "data", regexp = "Foreign Connected PAC")

pac <- read_csv(list_of_files, id = "year") |> 
  clean_names() |>
  mutate(year = str_extract(year, "\\d{4}(?=\\.csv$)") |> as.integer()) |>
  separate(country_of_origin_parent_company,
           into = c("country", "parent_company"),
           sep = "/", 
           fill = "right", 
           extra = "merge") |>
  mutate(across(c(dems, repubs), ~ as.numeric(str_replace_all(., "[\\$,]", "")))) |>
  select(year, pac_name_affiliate, country, parent_company, dems, repubs)

# Summarize totals by year for UK only
yearly_totals <- pac |>
  filter(country == "UK") |>
  pivot_longer(cols = c(dems, repubs), names_to = "party", values_to = "amount") |>
  filter(!is.na(amount)) |>
  group_by(year, party) |>
  summarise(amount = sum(amount), .groups = "drop")

# Plot
g2a <- ggplot(yearly_totals, aes(x = year, y = amount, color = party)) +
  geom_line(linewidth = 1.2) +     # Keep line only
  # geom_point(size = 2)      # Removed points
  
  # Removed geom_text() for labels (still commented out)
  # geom_text(aes(label = scales::dollar(amount, scale = 1e-6, suffix = "M")), 
  #           vjust = -0.5, size = 3.5, show.legend = FALSE) +
  
  scale_y_continuous(
    name = "Total amount",
    labels = label_dollar(scale = 1e-6, suffix = "M")
  ) +
  
  scale_x_continuous(
    name = "Year",
    breaks = scales::pretty_breaks()
  ) +
  
  scale_color_manual(
    values = c("dems" = "blue", "repubs" = "red"),
    labels = c("Democrats", "Republicans"),
    name = "Party"                          # - sets name on legend
  ) +
  
  labs(title = "Contributions to US political parties from UK-connected PACs") +
  
  theme(
    legend.position = c(0.9, 0.15),            # - getting a warning but if I do as requested, 
    axis.title.x = element_text(hjust = 0),    # - legend moves outside plot.
    axis.title.y = element_text(hjust = 0)
  )

print(g2a)

```

## 3 - Median housing prices in the US
<div style="background-color:#f0f0f0; padding:10px; border-radius:5px; border:1px solid #87CEEB;">
<b>Question #3</b><br>
<div style="background-color:#e0e0e0; padding:10px; border-radius:5px; border:1px solid #16DB65; display:flex; align-items:center; justify-content:center; height:60px;">
  <strong><br>Put the question here</strong>
</div>
</div>
```{r}
#| label: label-me-3
#| echo: false

```


## 4 - Expect More. Plot More.
<div style="background-color:#f0f0f0; padding:10px; border-radius:5px; border:1px solid #87CEEB;">
<b>Question #4</b><br>
<div style="background-color:#e0e0e0; padding:10px; border-radius:5px; border:1px solid #16DB65; display:flex; align-items:center; justify-content:center; height:60px;">
  <strong><br>Recreate the Target LOGO</strong>
</div>
</div>
```{r}
#| label: label-me-4
#| echo: false

library(ggplot2)
library(ggforce)  # for geom_circle()

# Define circles for the Target logo
circles <- data.frame(
  x0 = 0, y0 = 0,
  r = c(1, 0.7),
  fill = c("red", "white")
)

# Plot the logo + "TARGET" + ®
g1 <- ggplot() +
  # Draw the rings
  geom_circle(data = circles, aes(x0 = x0, y0 = y0, r = r, fill = fill), color = NA) + 
  # Solid center circle
  geom_point(aes(x = 0, y = 0), shape = 21, size = 30, fill = "red") +  
  # Add the word 'TARGET' below the logo
  annotate("text", x = 0, y = -1.4, label = "TARGET", color = "red", size = 10, fontface = "bold") +
  # Add the ® symbol offset to the bottom right of 'TARGET'
  annotate("text", x = 0.60, y = -1.50, label = "®", color = "red", size = 8) +
  # Style the plot
  scale_fill_identity() +
  coord_fixed() +
  theme_void()

plot(g1)


```
## 5 - Mirror, mirror on the wall, who’s the ugliest of them all?
<div style="background-color:#f0f0f0; padding:10px; border-radius:5px; border:1px solid #87CEEB;">
<b>Question #5</b><br>
<div style="background-color:#e0e0e0; padding:10px; border-radius:5px; border:1px solid #16DB65; display:flex; align-items:center; justify-content:center; height:60px;">
  <strong><br>Put the question here</strong>
</div>
</div>
```{r}
#| label: label-me-5
#| echo: false

```
